Malletwand: the Pendulum as a Handheld Interface to Musical Timing

### 2024-02-07 paper/paper.tex
(WIP) Paper

@@

Indeed, there have been a number of devices that manifest the link between periodic bodily movement and stable temporal patterns. Scrubbing is a standard technique for turntablists to manipulate the speed of sounds by rotating vinyl records. Hand-cranked musical boxes provide a more approachable way to drive the flow of music by one's own hands. The Swedish band Wintergatan went as far as to build an entire machine, the Marble Machine, that turns the revolution of the manual crank into the progression of a musical piece~\cite{Rundle_Woollaston-Webber_2017}.

@@

Our design adds to this ensemble a missing piece: a lightweight handheld controller that can be enjoyable and meaningful regardless of the participant's prior musical experience.

@@

@@

>>>>

%\begin{figure*}[h!]
%  \centering
%  \includegraphics[width=0.8 \textwidth]{placeholder.png}
%  \caption{The Malletwand controller with the prototype self-playing glockenspiel.}
%  \label{fig:Banner}
%\end{figure*}

@@

Indeed, there have been a number of devices that manifest the link between periodic bodily movement and stable temporal patterns. Scrubbing is a standard technique for turntablists to manipulate the speed of sounds by rotating vinyl records. Hand-cranked musical boxes provide a more approachable way to drive the flow of music by one's own hands; although the Swedish band Wintergatan went as far as to build an entire machine, the Marble Machine, that turns the revolution of the manual crank into the progression of a musical piece~\cite{Rundle_Woollaston-Webber_2017}.

@@

Still, these interfaces and instruments take larger physical forms and mainly aim at stage performances and/or dedicated experimental production setups. Our design adds to this ensemble a missing piece: a lightweight handheld controller interface that can be enjoyable and meaningful regardless of the participant's prior musical experience.

@@

\subsection{Design}
We first present the details of the handheld controller interface. It takes the form of a handle with a hanging bob (ball) that can freely swing and spin around, forming a pendulum. The participant swings the pendulum at their desired frequency while the bob's quasi-periodic motion is mapped to continuous musical time such that each fastest-moving point in a period corresponds to the start of a beat. (This point is usually perceived to be the ``central'' or the ``lowest'' point in simple pendulums, but since our pendulum has an extra degree of freedom, i.e., is a spherical pendulum, this appointment needs to be extrapolated.) Such generated timing signals are transmitted to an actuator (instrument) that plays a predetermined piece of music according to the manipulated flow of musical time.

Such controllers can

\subsection{Construction}
We design customised electronics for each of the parts -- the handle contains a battery, a vibration motor for haptic feedback, a radio transceiver, a microcontroller, and various peripheral components; the bob contains MEMS chips that form a 9-axis IMU (an accelerometer, a gyroscope, and a geomagnetometer). The complete structure is illustrated in figure~\ref{fig:WandConstruction}.

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=1\textwidth]{placeholder.png}
  \caption{(a) An illustration of the controller being operated by a participant. (b) A photo of the circuit board. (c) Block diagram.}
  \label{fig:WandConstruction}
\end{figure*}

We cut the board at the thin connection (in figure~\ref{fig:WandConstruction}a) and connect the parts with thin wires (8 strands of 32 AWG). The part for the handle is glued inside a 3D-printed casing with the wires hanging down to the bob. We add fishing wires to strengthen the connection and prevent the electric wires from excessive tension and fatigue.

\subsection{Signal processing}

@@

\subsection{Design}

\subsection{Construction}

\subsection{Performance}
